Learning constraints in spreadsheets and tabular data

### 2016-05-30 Paper/ECAI/tabular.tex
h!
Line spacing
worked with refs
some tweaking on related work

\usepackage[dvipsnames]{xcolor}
\newcommand{\sergey}[1]{\textcolor{magenta}{{\sc Sergey:} #1}\xspace}
\newcommand{\samuel}[1]{\textcolor{green}{{\sc Samuel:} #1}\xspace}
\newcommand{\tias}[1]{\textcolor{blue}{{\sc Tias:} #1}\xspace}
\newcommand{\luc}[1]{\textcolor{red}{{\sc Luc:} #1}\xspace}

@@

\begin{table*}
\caption{
  Overview of constraint templates implemented in \sname.
  Templates marked with $\dagger$ are \textit{aggregate} templates, the table shows the version for sum but \sname also supports max, min, average, product and count.
  Subgroups in normal font have to be vectors, whereas subgroups in \textbf{bold} may contain more than one vector.
  Templates marked with $*$ are \textit{structural}, i.e. they are not functional and cannot be implemented in most spreadsheet software.
  $\discrete(x)$ is a shortcut for: $\textual(x) \lor \numeric(x)$.
  {\small\luc{maybe say s.th. about some fancy constraints: example of a more complex one in the text (what does conditional mean, aggregate ... )}}
}
\label{table:constraints}
  {\centering
  \begin{tabularx}{\textwidth}{l X X}
    \textbf{\CName} & \textbf{\CSignature} & \textbf{\CFunction}\\ \hline \hline
    $\ecalldiff{\sg_x}^*$
      & $\discrete(\sg_x)$

      & All values in $\sg_x$ are different. $i \neq j$: $\sg_x[i] \neq \sg_x[j]$
      \\ \hline
    %X = Y & & \\

@@

      & $\sg_{x}[1] = 1$ and $\sg_{x}[i] = \sg_{x}[i - 1] + 1$.
      \\ \hline

@@

      & Every value in~$\sg_{fk}$ also exist in~$\sg_{pk}$ \\ \hline

@@

      & $\sbl{r}[i] = \sbl{val}[j]$ where $\sbl{pk}[j] = \sbl{fk}[i]$
      \\ \hline

@@

      & $\sbl{r}[i] = \sbl{val}[j]$ where $\sbl{pk}[j] \leq \sbl{fk}[i]$, $j$~maximal
      \\ \hline

@@

      & $\sbl{r}[i] = \sbl{1}[i] \times \eclookupf{\sbl{fk}}{\sbl{pk}}{\sbl{val}}{}[i]$.
      \\ \hline

@@

      & $\sg_{r}[i] = \sg_{1}[i] \times \sg_{2}[i]$.
      \\ \hline

@@

      & $\sg_{r}[i] = \sg_{1}[i] - \sg_{2}[i]$.
      \\ \hline

@@

      & At every position~$i$ in $1$ through $\plength(\sg_{r})$ there is exactly one vector~$v$ in $\mathbf{\sg_x}$ such that $v[i]$ is a non-blank value, then $v[i] = \sg_{r}[i]$.
      \\ \hline

@@

      & The values in $\sg_{r}$ represent the rank (from largest to smallest) of the values in $\sg_{x}$ (including ties)
      \\ \hline

@@

      & $\sg_{r}[i] = \sg_{r}[i - 1] + \sg_{pos}[i] - \sg_{neg}[i]$.
      \\ \hline

@@

      & $B_r[i] = \sum_{j = 1}^{\pcols(\bsbl{x})} \mathit{row}(i, \bsbl{x})[j]$
      \\ \hline

@@

      & $B_r[i] = \sum_{j = 1}^{\prows(\bsbl{x})} \mathit{column}(i, \bsbl{x})[j]$
      \\ \hline

@@

      & \[ B_r[i] = \sum_{j=1}^{\plength(B_{\textit{val}})} \begin{cases}
          \sg_{val}[j] & \text{if } \sg_{fk}[j] = \sg_{pk}[i] \\
          0 & \text{otherwise}
        \end{cases}
      \] \\ \hline

@@

\section{Related Work}\label{sec:related_work}
\sname combines ideas from several existing
approaches from multiple different fields of research.

First, it borrows techniques from logical and relational learning \cite{luc_book}, as it finds
constraints in multiple relations (or tables). However, unlike typical logical and relational learning
approaches, it focuses on data in spreadsheets and on constraints that can be column or row-wise.
Furthermore, it derives a set of simple “atomic” constraints rather than a set of rules that each consist of multiple literals as in clausal discovery \cite{claudien,lallouet}. Still several ideas from logical and relational learning
proved to be useful, such as the use of a canonical form, a refinement graph, and pruning for redundancy.
Also related to this line of research is the work on deriving constraints in databases such as functional and multi-valued dependencies \cite{savnik, Mannila-Raiha}
although this line of research has focused on more specialized techniques for specific types of constraints. Furthermore, the constraints used
in spreadsheets are often quite different than those in databases.

Second, there exist several algorithms in the constraint programming community that induce constraint (programs)
from one or more examples and questions. Two well-known lines of research include ModelSeeker \cite{modelseeker} and Quacq \cite{Quacq}.
The former starts from a single example in the form of a vector of values and then exhaustively looks for the most specific constraints from a large constraint library that hold in the vector. To this aim, it cleverly enumerates different tables from the vector. For instance, if the initial vector was of dimension $1 \times 20$, it would consider rearrangements of size $2 \times 10$, $4 \times 5$, $5 \times 4$, … and Modelseeker would then look for both column, row  and even diagonal constraints. Key differences with our technique are that we assume the tables are given (a reasonable assumption if one starts from a spreadsheet), and also that ModelSeeker’s constraints are the typical ones from the constraint programming community\tias{which is all over integers, we also over strings and real-valued, AND, not one vector but we have many indep. blocks}, which are aimed at combinatorial optimization and scheduling rather than what is typically used in spreadsheets. Quacq \cite{Quacq} and its predecessor Conacq \cite{Conacq} acquire a conjunction of constraints using techniques inspired on Mitchell's version spaces to process examples, some variants also asking informative questions to the user. As the conjunction for typical constraint programming problems (such as Sudoku) can be quite large, a key problem is to cope with redundant constraints\tias{it looks only at pairwise constraints, so an important factor is how fast the algorithm can converget to the entire constraint set. Redundant and spur also issue for these methods}. As Modelseeker, these systems focus on the types of constraints in combinatorial optimization but unlike ModelSeeker and our approach, they do not assume the variables are organized in tabular form.

Third, our work borrows from the seminal work of Gulwani et al. \cite{flashfill} on program synthesis for tabular data that is incorporated in Microsoft’s Excel. In FlashFill, the end-user might work on a table containing the surnames (column A) and names (column B) of politicians. The first row might contain the values A1 = ``Blair'' and B1 = ``Tony'', and the user might then enter in cell C1 the value ``T.B.''. At that point FlashFill would generate a program that produces the output ``T.B.'' for the inputs ``Blair'' and ``Tony'', would also apply it to automatically generate values for the other politicians (rows) in the table. While it might need an extra example to deal with names such as ``Van Rompuy, Herman'' and determine whether it should output ``H.V.'' or ``H.V.R.'', FlashFill learns correct programs from very few examples. Flashfill has been extended to e.g. FlashExtract \cite{flashextract} to produce a set of tables starting from a document (a txt, spreadsheet or HTML file) with a set of examples marked for extraction. However, unlike our approach, Flashfill requires the user to identify explicitly the desired inputs/outputs from the function as well as to provide explicitly some positive and sometimes also negative examples.  In contrast, our approach is unsupervised, although it would be interesting to study whether it could be improved through user interaction and through user examples.  Another interesting open issue (for all of these techniques with the exception of logical and relational learning) is how to deal with possibly incomplete or noisy data.

@@

Aside from finding errors, the system can be extended to deal with noise and the ability to learn soft constraints.
Soft constraints are (potentially weighted) constraints that hold only on some of the data.
This would extend the approach to new application domains as well as provide more native error detection.
\bibliographystyle{plain}
\bibliography{references}
\end{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\usepackage[dvipsnames]{xcolor}
\newcommand{\PADSEP}{-7pt}
\newcommand{\sergey}[1]{\textcolor{magenta}{{\sc Sergey:} #1}\xspace}
\newcommand{\samuel}[1]{\textcolor{green}{{\sc Samuel:} #1}\xspace}
\newcommand{\tias}[1]{\textcolor{blue}{{\sc Tias:} #1}\xspace}
\newcommand{\luc}[1]{\textcolor{red}{{\sc Luc:} #1}\xspace}

@@

\begin{table*}[!h]
\caption{
  Overview of constraint templates implemented in \sname.
  Templates marked with $\dagger$ are \textit{aggregate} templates, the table shows the version for sum but \sname also supports max, min, average, product and count.
  Subgroups in normal font have to be vectors, whereas subgroups in \textbf{bold} may contain more than one vector.
  Templates marked with $*$ are \textit{structural}, i.e. they are not functional and cannot be implemented in most spreadsheet software.
  $\discrete(x)$ is a shortcut for: $\textual(x) \lor \numeric(x)$.
  {\small\luc{maybe say s.th. about some fancy constraints: example of a more complex one in the text (what does conditional mean, aggregate ... )}}
}
\label{table:constraints}
  {\centering
  \begin{tabularx}{\textwidth}{l X X}
    \textbf{\CName} & \textbf{\CSignature} & \textbf{\CFunction}\\ \hline \hline
    $\ecalldiff{\sg_x}^*$
      & $\discrete(\sg_x)$

      & All values in $\sg_x$ are different. $i \neq j$: $\sg_x[i] \neq \sg_x[j]$
      \\[\PADSEP] \hline

@@

      & $\sg_{x}[1] = 1$ and $\sg_{x}[i] = \sg_{x}[i - 1] + 1$.
      \\[\PADSEP] \hline

@@

      & Every value in~$\sg_{fk}$ also exist in~$\sg_{pk}$ \\[\PADSEP] \hline

@@

      & $\sbl{r}[i] = \sbl{val}[j]$ where $\sbl{pk}[j] = \sbl{fk}[i]$
      \\[\PADSEP] \hline

@@

      & $\sbl{r}[i] = \sbl{val}[j]$ where $\sbl{pk}[j] \leq \sbl{fk}[i]$, $j$~maximal
      \\[\PADSEP] \hline

@@

      & $\sbl{r}[i] = \sbl{1}[i] \times \eclookupf{\sbl{fk}}{\sbl{pk}}{\sbl{val}}{}[i]$.
      \\[\PADSEP] \hline

@@

      & $\sg_{r}[i] = \sg_{1}[i] \times \sg_{2}[i]$.
      \\[\PADSEP] \hline

@@

      & $\sg_{r}[i] = \sg_{1}[i] - \sg_{2}[i]$.
      \\[\PADSEP] \hline

@@

      & At every position~$i$ in $1$ through $\plength(\sg_{r})$ there is exactly one vector~$v$ in $\mathbf{\sg_x}$ such that $v[i]$ is a non-blank value, then $v[i] = \sg_{r}[i]$.
      \\[\PADSEP] \hline

@@

      & The values in $\sg_{r}$ represent the rank (from largest to smallest) of the values in $\sg_{x}$ (including ties)
      \\[\PADSEP] \hline

@@

      & $\sg_{r}[i] = \sg_{r}[i - 1] + \sg_{pos}[i] - \sg_{neg}[i]$.
      \\[\PADSEP] \hline

@@

      & $B_r[i] = \sum_{j = 1}^{\pcols(\bsbl{x})} \mathit{row}(i, \bsbl{x})[j]$
      \\[\PADSEP] \hline

@@

      & $B_r[i] = \sum_{j = 1}^{\prows(\bsbl{x})} \mathit{column}(i, \bsbl{x})[j]$
      \\[\PADSEP] \hline

@@

      & \[ B_r[i] = \sum_{j=1}^{\plength(B_{\textit{val}})} \begin{cases}
          \sg_{val}[j] & \text{if } \sg_{fk}[j] = \sg_{pk}[i] \\
          0 & \text{otherwise}
        \end{cases}
      \] \\[\PADSEP] \hline

@@

\section{Related Work}\label{sec:related_work}
\sname combines ideas from several existing
approaches and different fields of research.

First, it borrows techniques from logical and relational learning \cite{luc_book}, as it finds
constraints in multiple relations (or tables). However, unlike typical logical and relational learning
approaches, it focuses on data in spreadsheets and constraints on columns or rows in the sheets.
Furthermore, it derives a set of simple “atomic” constraints rather than a set of rules that each consist of multiple literals as in clausal discovery \cite{claudien,lallouet}. Nevertheless, several ideas from logical and relational learning
proved to be useful, such as the use of a canonical form, a refinement graph, and pruning for redundancy.
Also related to this line of research is the work on deriving constraints in databases such as functional and multi-valued dependencies \cite{savnik, Mannila-Raiha}
although that line of research has focused on more specialized techniques for specific types of constraints. Furthermore, the constraints and formulas used
in spreadsheets are often different from those in databases.

Second, there exist a number of algorithms in the constraint programming community that induce constraint (programs)
from one or more examples and questions. Two well-known approaches include ModelSeeker \cite{modelseeker} and Quacq \cite{Quacq}.
The former starts from a single example in the form of a vector of values and then exhaustively looks for the most specific constraints from a large constraint library that hold in the vector. To this aim, it cleverly enumerates different de-vectorized tables; for instance, if the initial vector was of dimension $1 \times 20$, it would consider rearrangements of size $2 \times 10$, $4 \times 5$, $5 \times 4$, … and Modelseeker would then look for both column, row  and even diagonal constraints. Key differences with our technique are that we assume the tables are given (a reasonable assumption if one starts from a spreadsheet), and that ModelSeeker's constraints are the typical ones from the constraint programming community, aimed at combinatorial optimization and scheduling, and restricted to integer values; furthermore we can consider multiple tables and sheets. Conacq \cite{Conacq} acquires a conjunction of constraints using techniques inspired on Mitchell's version spaces to process examples; its successor Quacq uses active learning to query the user whether a partical constraint is satisfied or not. Each of the examples/queries is processed for each constraint, and hence an important issue is how fast the algorithm converges to the entire solution set (e.g. for Sudoku).
The focus is again on integer variables and typical combinatorial optimisation constraints.
%As the conjunction for typical constraint programming problems (such as Sudoku) can be quite large, a key problem is to cope with redundant constraints. As Modelseeker, these systems focus on the types of constraints in combinatorial optimization but unlike ModelSeeker and our approach, they do not assume the variables are organized in tabular form.

Third, our work borrows from the seminal work of Gulwani et al. \cite{flashfill} on program synthesis for tabular data, which is incorporated in Microsoft’s Excel. In FlashFill, the end-user might work on a table containing the surnames (column A) and names (column B) of politicians. The first row might contain the values A1 = ``Blair'' and B1 = ``Tony'', and the user might then enter in cell C1 the value ``T.B.''. At that point FlashFill generates a program that produces the output ``T.B.'' for the inputs ``Blair'' and ``Tony'', and also applies it to the other politicians (rows) in the table. While it might need an extra example to deal with names such as ``Van Rompuy, Herman'' and determine whether it should output ``H.V.'' or ``H.V.R.'', FlashFill learns correct programs from very few examples. Flashfill has been extended to e.g. FlashExtract \cite{flashextract} to produce a set of tables starting from an unstructured document (a txt or HTML file) with a set of examples marked for extraction. However, unlike our approach, Flashfill requires the user to identify explicitly the desired inputs/outputs from the function as well as to provide explicitly some positive and sometimes also negative examples.  In contrast, our approach is unsupervised, although it would be interesting to study whether it could be improved through user interaction and through user examples.  Another interesting open issue (for all of these techniques with the exception of logical and relational learning) is how to deal with possibly incomplete or noisy data.

@@

Aside from finding errors, the system can be extended to deal with noise and the ability to learn soft constraints.
Soft constraints are (potentially weighted) constraints that hold only on some of the data.
This would extend the approach to new application domains as well as provide more native error detection.

\bibliographystyle{plain}
{\footnotesize
  \bibliography{references}}
\end{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
